Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2018-07-30 submission/manuscript.Rmd
update introduction

Colorectal cancer (CRC) is currently the third highest cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment is attributed to being a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors are capable of being modified, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered [@Shah2017; @meta_analysis_crc_Sze2018]. Multiple reports in case/control studies have identified bacterial taxa commonly associated with individuals without tumors to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of these taxa within individuals wihtout tumors actively produce short-chain fatty acids (SCFAs) from fiber that are a part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Overall, the specific bacterial taxa of the microbiota that create SCFAs are an attractive target to modulate the risk of CRC.

Specific SCFAs, like butyrate, have shown positive results for CRC treatment within model systems [@test_OKeefe2016]. Butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects. For example, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus either patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. We also test previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity increases by increasing the number of patients with adenomas in our study. Additionally, we build upon these observations and assesses the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. Additionally, we (i) assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenoma or carcinoma using the Random Forest algorithm [@randomforest_citation_2002], (ii) used PICRUSt [@Langille2013] and metagenomic sequencing to assess the presence of genes involved in SCFA synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. This investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US [@Haggar2009; @Siegel2016]. Although there is a genetic component to the disease, the environment has been attributed to being a larger risk factor for CRC [@Lichtenstein2000]. These environmental risk factors include but are not limited to smoking cigarettes, diet, and the microbiota [@FlissIsakov2017; @Lee2015; @Kostic2011]. Many of these environmental risk factors, including the microbiota, could be easily modifiable, and this has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [@Zackular2013; @Baxter2014; @Zackular2015] and whether the bacterial community is altered in CRC [@Shah2017; @meta_analysis_crc_Sze2018]. Many of these previous case/control studies have identified resident bacterial taxa to be decreased in patients with carcinoma tumors [@Zeller2014; @Baxter2016; @meta_analysis_crc_Sze2018]. Many of the bacterial species from these resident taxa actively produce short-chain fatty acids (SCFAs) from fiber that are part of our general diet [@Holscher2017]. The most extensively studied of these SCFAs are acetate, butyrate, and propionate [@Louis2016]. Ultimately, these SCFAs could be the main metabolites involved with protection against tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown positive results within model systems [@test_OKeefe2016]. For example, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems [@Encarnao2018]. Additionally, fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while also increasing SCFA concentrations [@Bishehsari2018]. These exciting results in model systems suggests that supplementation with food sources that bacteria use to create these SCFAs may be able to confer beneficial effects against CRC. 
However, it is important to note that these model systems provide only preliminary evidence towards the ability of SCFAs to reduce and treat tumors and the studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas or individuals without colon tumors [@Ohigashi2013]. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence [@Yao2017]. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC, there is a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it tests previous case/control findings on SCFA concentrations in individuals with and without tumors. It also tests previous suggestions that there is a continuous reduction in SCFA concentrations as tumor severity worsens by increasing the number of patients with adenomas within the study. Additionally, we build upon these observations by assessing the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a risk stratification tool of colorectal tumors (adenoma or carcinoma). We also investigate whether OTUs that are most important to these models are closely associated with the classification of SCFA concentrations. Collectively, this study provides important information on the replicability of previous findings in humans by extensively studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group had a sample obtained at a single cross sectional point in time while the second group had samples obtained before (pre-) and after (post-)treatment for colorectal tumors. We also used PICRUSt [@Langille2013] and metagenomic sequencing to investigate if there were any differences in genes involved with SCFA synthesis between individuals without tumors, patients with adenomas, and patients with carcinomas. Next, using the cross-sectional data, we analyzed the number of correlations between OTU relative abundance and SCFA concentrations across individuals without tumors and patients with adenomas or carcinomas. Additionally, we assessed the affect adding SCFA concentrations to OTU data had on classification of patients with adenomas or carcinomas using the Random Forest algorithm [@randomforest_citation_2002]. We also analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations. Collectively, this investigation provides additional information as to whether SCFAs are decreased in patients with colorectal tumors and provides context as to whether targeting taxa to increase SCFA concentrations is a viable option to protect against colon tumorigenesis.

### 2018-07-30 submission/manuscript.tex
update introduction

Colorectal cancer (CRC) is currently the third highest cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment is attributed to being a
larger risk factor for CRC (3). These environmental risk factors include
but are not limited to smoking cigarettes, diet, and the microbiota
(4--6). Many of these environmental risk factors are capable of being
modified, and this has lead to the investigation of how the microbiota
may exacerbate or cause tumorigensis (7--9) and whether the bacterial
community is altered (10, 11). Multiple reports in case/control studies
have identified bacterial taxa commonly associated with individuals
without tumors to be decreased in patients with carcinoma tumors
(11--13). Many of these taxa within individuals wihtout tumors actively
produce short-chain fatty acids (SCFAs) from fiber that are a part of
our general diet (14). The most extensively studied of these SCFAs are
acetate, butyrate, and propionate (15). Overall, the specific bacterial
taxa of the microbiota that create SCFAs are an attractive target to
modulate the risk of CRC.

Specific SCFAs, like butyrate, have shown positive results for CRC
treatment within model systems (16). Butyrate has been shown to inhibit
cancer cell growth in \emph{in vitro} systems (17). Additionally,
supplementation with food sources that bacteria use to create these
SCFAs may also be able to confer beneficial effects. For example, fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (18). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus either patients with adenomas or
individuals without colon tumors (19). Although this would argue that
increasing SCFA concentrations could be protective against
tumorigenesis, fiber supplementation in randomized controlled trials
have consistently failed to protect against tumor recurrence (20). These
findings would argue against the utility of treatments that aim to use
SCFAs to reduce or protect against tumorigenesis. Given the lack of
clear evidence in human studies of the benefit of SCFAs in CRC, there is
a need for more investigation into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, it
tests previous case/control findings on SCFA concentrations in
individuals with and without tumors. We also test previous suggestions
that there is a continuous reduction in SCFA concentrations as tumor
severity increases by increasing the number of patients with adenomas in
our study. Additionally, we build upon these observations and assesses
the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a
risk stratification tool of colorectal tumors (adenoma or carcinoma).
Collectively, this study provides important information on the
replicability of previous findings in humans by extensively studying how
SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we (i) assessed the affect adding SCFA concentrations to
OTU data had on classification of patients with adenoma or carcinoma
using the Random Forest algorithm (21), (ii) used PICRUSt (22) and
metagenomic sequencing to assess the presence of genes involved in SCFA
synthesis, and (iii) analyzed how well 16S rRNA gene sequencing predicts
SCFA concentrations. This investigation provides additional information
as to whether SCFAs are decreased in patients with colorectal tumors and
provides context as to whether targeting taxa to increase SCFA
concentrations is a viable option to protect against colon
tumorigenesis.

@@

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (22) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

@@

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification and regression models via the
caret package (21, 43). This was repeated on 100 different 80/20 splits
of the data to generate a reasonable range for the AUC of the model. The
reported AUCs, unless otherwise specified, are for the test sets. The
classification models were built to group normal versus adenoma, normal
versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status.

@@

\hypertarget{ref-randomforest_citation_2002}{}
21. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\hypertarget{ref-Langille2013}{}
22. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the environment has been attributed to being a
larger risk factor for CRC (3). These environmental risk factors include
but are not limited to smoking cigarettes, diet, and the microbiota
(4--6). Many of these environmental risk factors, including the
microbiota, could be easily modifiable, and this has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(7--9) and whether the bacterial community is altered in CRC (10, 11).
Many of these previous case/control studies have identified resident
bacterial taxa to be decreased in patients with carcinoma tumors
(11--13). Many of the bacterial species from these resident taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are
part of our general diet (14). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (15). Ultimately, these
SCFAs could be the main metabolites involved with protection against
tumorigenesis and could help to reduce the risk of CRC.

Prior research suggests that SCFAs have promise in acting as an
anti-tumorigenic agent. Specific SCFAs, like butyrate, have shown
positive results within model systems (16). For example, butyrate has
been shown to inhibit cancer cell growth in \emph{in vitro} systems
(17). Additionally, fiber supplementation in mouse models of CRC caused
an overall reduction in tumor burden while also increasing SCFA
concentrations (18). These exciting results in model systems suggests
that supplementation with food sources that bacteria use to create these
SCFAs may be able to confer beneficial effects against CRC. However, it
is important to note that these model systems provide only preliminary
evidence towards the ability of SCFAs to reduce and treat tumors and the
studies reporting benefit in humans has been less convincing.

Overall, there is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas or individuals
without colon tumors (19). Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (20). These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC, there is a need for more
investigation into this area.

Our study fills some of the current gaps in the literature that relate
to the study of SCFAs and CRC in human populations. Specifically, it
tests previous case/control findings on SCFA concentrations in
individuals with and without tumors. It also tests previous suggestions
that there is a continuous reduction in SCFA concentrations as tumor
severity worsens by increasing the number of patients with adenomas
within the study. Additionally, we build upon these observations by
assessing the utility of using SCFAs and Operational Taxonomic Units
(OTUs) as a risk stratification tool of colorectal tumors (adenoma or
carcinoma). We also investigate whether OTUs that are most important to
these models are closely associated with the classification of SCFA
concentrations. Collectively, this study provides important information
on the replicability of previous findings in humans by extensively
studying how SCFAs are associated with colorectal tumors.

To accomplish this task we directly measured the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors. We also
used PICRUSt (21) and metagenomic sequencing to investigate if there
were any differences in genes involved with SCFA synthesis between
individuals without tumors, patients with adenomas, and patients with
carcinomas. Next, using the cross-sectional data, we analyzed the number
of correlations between OTU relative abundance and SCFA concentrations
across individuals without tumors and patients with adenomas or
carcinomas. Additionally, we assessed the affect adding SCFA
concentrations to OTU data had on classification of patients with
adenomas or carcinomas using the Random Forest algorithm (22). We also
analyzed how well 16S rRNA gene sequencing predicts SCFA concentrations.
Collectively, this investigation provides additional information as to
whether SCFAs are decreased in patients with colorectal tumors and
provides context as to whether targeting taxa to increase SCFA
concentrations is a viable option to protect against colon
tumorigenesis.

@@

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (21) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formatted table using the biom convert function, the subsequent biom
file was processed with the `normalize\_by\_copy\_number.py' function,
and subsequent imputed metagenomes created using the
`predict\_metagenomes.py' function.

@@

\textbf{Random Forest models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for classification and regression models via the
caret package (22, 43). This was repeated on 100 different 80/20 splits
of the data to generate a reasonable range for the AUC of the model. The
reported AUCs, unless otherwise specified, are for the test sets. The
classification models were built to group normal versus adenoma, normal
versus carcinoma, and high versus low SCFA concentrations. The
regression models were built to classify the SCFA concentrations of
acetate, butyrate, and propionate regardless of disease status.

@@

\hypertarget{ref-Langille2013}{}
21. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-randomforest_citation_2002}{}
22. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.
